---
title: "Introduction to R Biostatistics Analysis"
author: "Gabriel Demetrios Lafis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R Biostatistics Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rbiostatistics)
```

## Introduction

The `rbiostatistics` package provides a comprehensive set of tools for biostatistical analysis, with a focus on survival analysis and clinical trials. This vignette introduces the main functionality of the package and demonstrates how to use it with example data.

## Survival Analysis

Survival analysis is a branch of statistics that deals with the analysis of time-to-event data. The `rbiostatistics` package provides several functions for survival analysis:

### Creating a Survival Object

The first step in survival analysis is to create a survival object:

```{r}
# Load sample data
data(lung_cancer_survival)

# Create a survival object
surv_obj <- create_survival_object(lung_cancer_survival, 
                                  time_col = "time", 
                                  status_col = "status")
head(surv_obj)
```

### Kaplan-Meier Survival Curves

The Kaplan-Meier method is a non-parametric statistic used to estimate the survival function from lifetime data:

```{r}
# Fit Kaplan-Meier curves by sex
km_fit <- fit_kaplan_meier(lung_cancer_survival, 
                          time_col = "time", 
                          status_col = "status", 
                          group_col = "sex")

# Plot the Kaplan-Meier curves
plot(km_fit, 
     xlab = "Time (months)", 
     ylab = "Survival Probability",
     main = "Kaplan-Meier Survival Curves by Sex",
     col = c("blue", "red"))
legend("topright", 
       legend = c("Male", "Female"), 
       col = c("blue", "red"), 
       lty = 1)
```

### Log-Rank Test

The log-rank test is used to compare the survival distributions of two or more groups:

```{r}
# Perform log-rank test
log_rank_result <- log_rank_test(lung_cancer_survival, 
                                time_col = "time", 
                                status_col = "status", 
                                group_col = "sex")
print(log_rank_result)
```

### Cox Proportional Hazards Model

The Cox proportional hazards model is a regression model commonly used in survival analysis:

```{r}
# Fit Cox model
cox_model <- fit_cox_model(lung_cancer_survival, 
                          time_col = "time", 
                          status_col = "status", 
                          covariates = c("age", "sex", "ph.ecog", "stage"))

# Print model summary
summary(cox_model)

# Calculate hazard ratios
hr <- calculate_hazard_ratio(cox_model)
print(hr)
```

## Clinical Trials Analysis

Clinical trials are research studies performed to evaluate the safety and efficacy of medical interventions. The `rbiostatistics` package provides several functions for clinical trials analysis:

### Sample Size Calculation

Before conducting a clinical trial, it's important to determine the appropriate sample size:

```{r}
# Calculate sample size for a two-group comparison
n <- calculate_sample_size(effect_size = 0.5, power = 0.8, alpha = 0.05)
print(paste("Required sample size per group:", ceiling(n)))
```

### Randomization

Randomization is a crucial step in clinical trials to minimize bias:

```{r}
# Create a sample of 20 subjects
subjects <- data.frame(id = 1:20)

# Randomize subjects to treatment and placebo groups
randomized <- randomize_subjects(subjects, 
                                groups = c("Treatment", "Placebo"), 
                                ratio = c(1, 1))
table(randomized$group)
```

### Treatment Effect Analysis

After collecting data, we can analyze the treatment effect:

```{r}
# Load sample data
data(clinical_trial_data)

# Analyze treatment effect
result <- analyze_treatment_effect(clinical_trial_data, 
                                  outcome = "week12", 
                                  treatment = "treatment", 
                                  covariates = c("baseline", "age", "sex"))
print(result)
```

### Subgroup Analysis

Subgroup analysis can help identify whether the treatment effect varies across different patient populations:

```{r}
# Analyze treatment effect in subgroups
subgroup_results <- analyze_subgroups(clinical_trial_data, 
                                     outcome = "week12", 
                                     treatment = "treatment", 
                                     subgroups = c("sex", "site"))
print(subgroup_results)
```

## Conclusion

The `rbiostatistics` package provides a comprehensive set of tools for biostatistical analysis, with a focus on survival analysis and clinical trials. This vignette has introduced the main functionality of the package and demonstrated how to use it with example data.

For more information, please refer to the function documentation and other vignettes.

